% ----------------------------------------------------------------
% AMS-LaTeX Paper ************************************************
% **** -----------------------------------------------------------
\documentclass[oneside]{amsart}
\usepackage{graphicx}
\usepackage{color}
\usepackage[letterpaper]{geometry}
\usepackage[colorlinks=false,
            pdfborder={0 0 0},
            pdftitle={CSC488 A5},
            pdfauthor={Daniel Bloemendal},
            pdfsubject={CSC488},
            pdfstartview=FitH,
            pdfmenubar=false,
            pdfdisplaydoctitle=true,
            bookmarks=false]{hyperref}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{mathtools}
\usepackage{listings}
\usepackage{float}
\usepackage[table]{xcolor}
% ----------------------------------------------------------------
\vfuzz2pt % Don't report over-full v-boxes if over-edge is small
\hfuzz2pt % Don't report over-full h-boxes if over-edge is small
% THEOREMS -------------------------------------------------------
\newtheorem{thm}{Theorem}[section]
\newtheorem{cor}[thm]{Corollary}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{prop}[thm]{Proposition}
\theoremstyle{definition}
\newtheorem{defn}[thm]{Definition}
\theoremstyle{remark}
\newtheorem{rem}[thm]{Remark}
\numberwithin{equation}{section}
% MATH -----------------------------------------------------------
\newcommand{\norm}[1]{\left\Vert#1\right\Vert}
\newcommand{\abs}[1]{\left\vert#1\right\vert}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\Real}{\mathbb R}
\newcommand{\eps}{\varepsilon}
\newcommand{\To}{\longrightarrow}
\newcommand{\BX}{\mathbf{B}(X)}
\newcommand{\A}{\mathcal{A}}
\newcommand{\e}{\mathrm{e}}
\newcommand{\AND}{\wedge}
\newcommand{\OR}{\vee}
\newcommand{\NOT}{\neg}
\newcommand{\IMPLIES}{\to}
\newcommand{\TRUE}{\top}
\newcommand{\FALSE}{\bot}
\newcommand{\EQUALS}{\equiv}
\DeclareMathOperator{\sech}{sech}
\newcolumntype{B}{>{\columncolor{black}\color{white}}c}
% ----------------------------------------------------------------
\lstset {
    frame=top,
    frame=bottom,
    numbers=none,
    xleftmargin=14pt,
    framexleftmargin=17pt,
    framexrightmargin=17pt,
    framexbottommargin=2pt,
    framextopmargin=2pt
}
\captionsetup[lstlisting]{singlelinecheck=false, margin=0pt, font={sf},labelsep=space,labelfont=bf}
% ----------------------------------------------------------------

\begin{document}

\title[CSC488 A5]{CSC488\\ASSIGNMENT 5\\Code Generator}
\author{Daniel Bloemendal \\ Oren Watson}

% ----------------------------------------------------------------
\begin{titlepage}
\maketitle
\thispagestyle{empty}
\tableofcontents
\end{titlepage}
% ----------------------------------------------------------------

\section{Instructions}
Before delving into the design of the code generator, it deserves noting that there are a couple of
features that go somewhat beyond the assignment requirements and are therefore disabled by default.
We will start with a brief discussion of the first of the two, bounds checking.

\subsection{Bounds checking}
Disabled by default, the code generator has the ability to add two different levels of array bounds
checking. The first level is ``simple'' bounds checking. It can be enabled with the compiler flag
``\texttt{-B s}''. In both modes, array bounds are checked on array access. However, in the simple
mode when an error occurs a simple error message with limited location information is provided. The
second level is ``enhanced'' bounds checking, enabled with the compiler flag ``\texttt{-B e}''. It
differs from the second mode in the level of detail provided in the location information contained
in the error message. An example of an out of bounds error message is provided below.
\begin{lstlisting}[caption=Out of bounds error, basicstyle=\fontsize{7}{10}\selectfont\ttfamily]
db@vmbox ~/workspace/CSC488/A5 $ ./RUNCOMPILER.sh -B e testing/fail/runtime_bounds2d.488
...
Start Execution  pc = 0, msp = 311, mlp = 16383

testing/fail/runtime_bounds2d.488:7:5: subscript out of range
    a[1, 11] := 1
         ^^
...
\end{lstlisting}
One of the reasons bounds checking is disabled by default is due to the tendency of the feature to
greatly increase the generated program size. This is mostly due to the emitted text strings in the
constant string pool. Simple bounds checking reduces this problem somewhat, as the text strings are
less complex and varied. Since identical strings resolve to the same address in the text constant
section, this reduces program size a fair bit.

\subsection{Code dump \& syntax highlighting}
The second of the two features is syntax highlighting for dumped code. The code generator can be
requested to dump code in two ways. The command line option ``\texttt{-D x}'' will prompt the
generator to dump the generated code. However, the machine will also dump the final disassembly of
the program. For only the IR assembly output use the command line option ``\texttt{-T c}''. To add
syntax highlighting to the dump, coloured via ANSI colour codes, add the command line option
``\texttt{--syntax-highlighting}''.

\section{Design}
\subsection{Overview}
The overarching theme in the design of the code generator was to avoid exposing the code generator
class \texttt{CodeGen} to the complexities and finer details of the underlying machine. To that end,
an assembler was developed that hides the complexity of addressing code via a label system and
provides an enhanced instruction set, simplifying the emitted code in \texttt{CodeGen}. It should be
noted that the assembler is also entirely decoupled from the rest of the code generator and stands
on its own. The assembler is covered in more detail in \texttt{doc/ASSEMBLER.pdf}. We made a
distinction  between `major' and `minor' scopes. A major scope is a whole program, a function, or a
procedure, requiring its own stack frame; all other scopes are considered minor scopes. The
complexities of  managing major scopes, their displays, and ensuring that minor scopes are merged
into their enclosing  major scopes, are dealt with by the \texttt{Frame} and \texttt{Table} classes.

\subsection{Coalescing minor scopes}
One of the more involved parts of developing the code generator was to ensure that minor scopes were
coalesced into their enclosing major scope. This is rather important as it avoids draining the
rather limited resource of display registers, 16 in total. In addition, it reduces the size of the
generated code as it avoids having to generate a prolog and epilog for every minor scope. The main
class responsible for managing this is \texttt{Frame}. It lays out the combined locals of all minor
scopes in a given major scope. It does so via an algorithm identical to the record layout algorithm
discussed in class. A frame can be thought of as a record corresponding to a major scope where all
minor scopes are sub-records. One important detail is that sibling minor scopes can be thought of as
belonging to a union. The reason this is a reasonable thing to do is because only one sibling minor
scope is alive at any one time. Therefore, it is safe to overwrite the locals used by any previously
executed sibling minor scope.

We will now proceed to an example. We will define a small program and lay out the locals according
to the algorithm in \texttt{Frame}.
\begin{lstlisting}[caption=Major scope in 488]
{
    var a : integer a := 0
    { var b, c : integer b := 0 c := 0 }
    { var d, e, f : integer d := 0 e := 0 f := 0 }
    { var g : integer g := 0
        { var h : integer h := 0 }
        { var i, j : integer i := 0 j := 0 }
    }
}
\end{lstlisting}

\noindent We will proceed to lay out the locals. We will express our layout using a C structure.
\begin{lstlisting}[caption=Layout as C structure, language=C, directivestyle={\color{black}} emph={short}, emphstyle={\color{blue}}]
struct major_1 {
    short a;
    union {
        struct { short b, c; } minor_1;
        struct { short d, e, f; } minor_2;
        struct {
            short g;
            union {
                struct { short h; } minor_4;
                struct { short i, j; } minor_5;
            } _siblings;
        } minor_3;
    } _siblings;
};
\end{lstlisting}

\subsection{Calling convention}
A key issue that we debated at length was the method by which function and procedure calls would be
implemented. We decided that the most important thing in functions was for the calling code  to have
the result value at the top of the stack when it returns, with no cleanup, so that the result could
immediately be worked with. Therefore, the function cleans up its own arguments and places its
result into a reserved place at the bottom of the stack frame. We also decided that to avoid undue
complexity, we would treat procedures identically to functions, even if this resulted in wasting
space with an unused result value space.

To call a function, the calling code reserves a place for the result, pushes the return address and
arguments, and then jumps to the function code. The function code then saves the display pointer for
its level, sets the display, and allocates space for its locals. When returning, the function
deallocates its locals, restores the display pointer to its previous state, pops its arguments off
the stack, and finally returns, leaving the result as the only thing left. The frame therefore has
the following structure, shown in figure \ref{fig:frame}, with the stack growing down. It should be
noted that the base address $D[LL]$ refers to the display set during the prolog of a function via
the \texttt{SAVECTX LL} intermediate instruction. Here, $LL$ refers to the lexical level of the
major scope.

\begin{figure}[h]
\caption{Stack frame}
\begin{tabular}{l|l}
Result & $D[LL] - N - 3$ \\
Return address & $D[LL] - N - 2$  \\
Argument $1$ & $D[LL] - N - 1$  \\
... & ... \\
Argument $N$ & $D[LL] - 2$ \\
Previous display & $D[LL] - 1$ \\
Locals & $D[LL]$ \\
\end{tabular}
\label{fig:frame}
\end{figure}

\subsection{Error handling}
\subsubsection{Integer size}
\subsubsection{String size}
\subsubsection{Array size}
\subsubsection{Locals size}

% ----------------------------------------------------------------
\end{document}
% ----------------------------------------------------------------
